---
title: "01 Raster Basics"
author: "G.Perkins (Ministry of Environment and Climate Change Strategy)"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE

---
<!--
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, include=FALSE}
knitr::opts_knit$set(echo = TRUE)

```


This document outlines some basic operations for rasters. You can follow along with the HTML or run the code yourself.  

```{r, eval = FALSE, echo = FALSE}
library(usethis)

# 1)  if you have not already - download the course content: 
 use_course("bcgov/bcgov-r-geo-workshop")

# 2) OPTIONAL : to generate an R script with code only

library(knitr)

purl("20191106_Day_2_PM_Raster/01_Raster_Basics.Rmd", 
    output = "20191106_Day_2_PM_Raster/01_Raster_Basics_code.R") 

```


### Package history and tools

- `raster` (first released in 2010 Robert Hijmans)
- `sp` including  `SpatialGridDataFrame` & `SpatialPixalsDataFrame` data structure
- `stars` : Spatiotemporal Arrays, Raster and Vector Data Cubes (Edzer Pebesma).
- helper packages : `fasterize`, `rgdal`, `mapview` 
- tools and bridges (RSAGA, RQGIS, RPYGeon)


This modual will primarily be usng the [raster](https://cran.r-project.org/web/packages/raster/index.html) package, however it is worth noting there are a number of other packages currently being developed to handle raster data efficiently. These include; [stars](https://cran.r-project.org/web/packages/stars/index.html) and [terra](https://www.rdocumentation.org/packages/terra/versions/0.3-7/topics/terra-package).  


### 1. Set up 
Firstly we set up our session by reading in required libraries and dataset. We can use `list.files` to take a look at what file types are available.  

```{r session setup, echo = TRUE, results = "hide", message= FALSE}

library(raster)

# set up the data location --------------------------------------

zip.file <- "../data/20191106_Day_2_PM_Raster/raster_basics.zip"
data.dir <-"../data/20191106_Day_2_PM_Raster/raster_basics"

unzip(zipfile = zip.file, exdir = data.dir)

```


```{r optional local drive, echo = FALSE, eval = FALSE }
# optional: To run locally within Rmarkdown 

data.dir <-"data/20191106_Day_2_PM_Raster/raster_basics/raster_basics/"

list.files(data.dir)

```


### 2. Loading and checking rasters 

We can read in raster data using the `raster` package. Unlike vectors, Rasters are defined by the location of grid extents, rather than individual vertices. This has some advantages and disadvantages. 


```{r load Raster, echo = TRUE, message= FALSE, include = T, results = "hide"}

# read in a raster 

file = list.files(data.dir, "DEM.tif", recursive = T, full.names = T)

dem <- raster(file)

```

Once we have read in the raster we can look at the key information; dimension (`dem`), resolution (cell size)(`res`), extents (`extent`), and Coordinate Reference System (CRS).

```{r, echo = T, include = T,eval = FALSE}

# dimension 
dim(dem)

#resolution 
res(dem)

# extent
extent(dem)

```

Secondly we can look in detail at the data values. For example we can review the minimum, maximum values and spread of the data (ie. the distribution of values)

```{r, echo = TRUE, include = TRUE, message = FALSE, eval = FALSE}

summary(dem) # Note this throws an error 
 
summary(dem, maxsamp = ncell(dem)) # forces to check all raster 

cellStats(dem, sd)

minValue(dem)

maxValue(dem)

```

As with any `r` object we can do a quick check of the datasets using standard functions (`head`). 

```{r, explore the data , echo = T, include = TRUE, eval = FALSE}

head(dem)           # not very useful

values(dem)         # a little more useful 

unique(values(dem)) # much more useful 

```

We also want to take a look to see if the raster data is what we expect. We can do this with a simple `plot()` function. 

```{r basic plot, results = "hide"}

plot(dem)

```

When reading in rasters it is also important to check how NA's are treated. In this case our `NA`'s are correctly assigned as we see the min and max values do not include NA's. If we read in a raster as see `NA`'s in the min and max values summary we can assign `NA` or reassign to a specific value. For example we can assign all `NA` values a value of 0 or -9999.


```{r, include = TRUE, results = "hide"}
dem <- setMinMax(dem)  

NAvalue(dem) = -9999

```

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

#### _Questions:_ 

1) Given the code below: what is the extent, cell size and maximum value of the raster ? 

```{r, eval = T, echo = FALSE}
sfile = list.files(data.dir, "slope.tif", recursive = T, full.names = T)
slope <- raster(sfile)
slope

```

2) What code could we use to check the mean values of all cells in the raster? 

</div>


### 3. Coordinate Reference System (CRS)

Coordinate Reference Systems come in three formats (`ESPG`, `proj4string`, `WKT`). Raster data commonly relies on `proj4string` type. In this case the `+` separates each element. You can think of this as a common separating each value in a csv file. Tip: Mapping with leaflet and other web-mapping requires WGS84 (`EPSG: 4326`). We can see the crs when we call the raster by name or directly using the `crs()` command.

```{r projections, include = TRUE}
dem
crs(dem)

```

The `crs` gives us important information as to what units the data is in, and as a consequence the cell size. For example our `dem` raster is measured in meters (m). This means out resolution is 25 m x 25 m cell size. 
This is important to understand when looking at area calculations and cell resolution. 

In many cases we use BCAlbers (`EPSG:3005`) as it is an equal area projection with units in meters. 

```{r, include = TRUE, results = "hide"}
# we can check the information on our ESPG:3005
CRS("+init=epsg:3005")

projection(dem)  # gives us the string version proj4string

```


#### Reprojecting Rasters

Reprojecting is different between rasters and vectors. In vectors, each coordinate is projected separately, however as rasters are based on a grid of cells the same size this method is not possible. Raster projections involves creating a new raster with the desired resolution. Each cell is then estimated or assigned a new value. Note the number of columns and rows are likely different from the original raster. 

We can use the `projectRaster()` function. This require a `raster` object and a `crs` argument. This function only accepts `proj4string` definitions of a CRS rather than EPSG codes directly. 


```{r reproject rasters, include = TRUE, eval = FALSE}
crs(dem)

wgs84 <- "+init=epsg:4326"

dem.wgs84 = projectRaster(dem, crs = wgs84)

plot(dem)
plot(dem.wgs84)

nrow(dem)
nrow(dem.wgs84)

res(dem)
res(dem.wgs84)

```

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

#### _Questions:_ 

1) Compare the extent on the`dem` and `dem.wgs84` ? 

2) Reproject the `dem.wgs84` raster back to `EPGS:3005` and check the crs to confirm ?

</div>


### 4. Visualise Rasters and Raster Data

We can do quick visualisations of our rasters using the `plot()`. We can use `mapview` or `rasterVis` to view the plot interactively. Another option to take advantage of the `ggplot2` package is to covert our raster into a data.frame and plot the xy variables. This however can be more time consuming as we see from the code below.


```{r raster viz, include = TRUE, eval = FALSE, warning= FALSE,  message = FALSE}
# Option 1: quick check of the raster 
plot(dem)

# Option 2: interactive mapping 
library(mapview)
mapview(dem)

# Option 3: more fancy exploring 
library(rasterVis)
# plot3D(dem)

# Option 4: convert to data frame and plot 
library(ggplot2)
dem_df <- as.data.frame(dem, xy = TRUE)
str(dem_df)

ggplot() +
    geom_raster(data = dem_df , aes(x = x, y = y, fill = DEM)) +
    scale_fill_viridis_c() +
    coord_quickmap()

```

We can also do a plot of the raster values. For example we can create a histogram to look for any anomolies.

```{r histogram, include = TRUE, eval = FALSE}

histogram(dem)

```


### 5. Raster Calculations 

We can do very fast calculations on raster objects as they do not contain explicit spatial projections. We can perform calculations based at difference scales; Local (per cell), Focal (neighbourhood), Zonal (neighburhood with iregular size and shape), Global (per raster calculations). 

```{r, eval = FALSE }

# Raster Calculations - per cell 
dem + dem 

log(dem)

```

We can reclassify values within the raster. This is useful if we want to bin values into catergories or define a subset of data. Using our `dem` raster we can reclassify our values into groups by creating a matrix with defined range and assigned value. We know our minimum value is `r minValue(dem)` and maximum value is `r maxValue(dem)`. 

```{r reclass raster, results = 'hide'}
# reclasss a raster to above or below 1000

rcl <- matrix(c(0, 1000, 1, 1000.0001, 1600, 2), 
              ncol = 3, 
              byrow = TRUE)
rcl 

recl.dem <- reclassify(dem, rcl = rcl)
plot(recl.dem)

```

Focal operations take into account a cell and its neighbours. Typically these include a 3 x 3 cells. This operation applies an aggregate function to all cells within the neighbourhood with the result being the central cell value. 

We need to define the shape of a moving window with a matrix with corresponding weights. This is commonly summary functions, ie `sum()`, `mean()`, `var()`.

```{r focal , eval = FALSE, include = TRUE}
#Focal operations

r_focal = focal(dem, w = matrix (1, nrow = 3, ncol = 3), fun = min)

plot(dem)
plot(r_focal)

```


Zonal statistics allows us to calculate summary statistics per data type or number. For example we can calculating mean, max values. The output is in tabular format. As a demonstration we can use out reclassed raster from above as a quick demonstration. 

```{r zonal stats demo, eval = FALSE}
#zonal calculations 

plot(recl.dem) 

dem.zone <- zonal(dem, recl.dem, fun= "mean")

dem.zone

```

We will talk more about zonal statistics for remote sensed datasets. 

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">


#### _Questions_

1) Reclasify the dem raster into four classes ( <800m, 800 - 1000, 1000-1250, 1250 - maximum) and plot the reclassed raster.  

2) Calculate the standard deviation of the dem values in each zone you created. Which zone has the most variation? 


3) Create a map shpwing variance over a 9 x 9 neighbourhood.


2) Calculate the mean value for the entire raster (hint: see section 1.)


</div>
<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">

#### _Bonus Section_

We can use focal operations to generate slope and aspect from our base dem raster. For this we will use the `terrain()` function. Run through this code and plot the outputs. 

```{r create a hillshade map, eval = T , results="hide"}

slope <- terrain(dem, opt='slope', unit='radians',  neighbors=8)

aspect <- terrain(dem, opt='aspect', unit='radians',  neighbors=8)


```

We can now use these outputs to create a hillshade. 

```{r hillshade, results = "hide"} 

hs <- hillShade(slope, aspect, angle=30, direction=270)
plot(hs, col=gray.colors(20))

```

You can adjust the angle and direction above to see the influence it has on the hillshade output.  

We can plot hillshade and dem to create nice feature maps.  

```{r ggplot mapping, eval = FALSE}
hs_df <- as.data.frame(hs, xy = TRUE) 
dem_df <- as.data.frame(dem, xy = TRUE)
  
ggplot() +
  geom_raster(data = dem_df , 
              aes(x = x, y = y, 
                  fill = DEM)) + 
  geom_raster(data = hs_df,
              aes(x = x, y = y, alpha = layer)) + 
  scale_fill_gradientn(colours = terrain.colors(10)) +  
  scale_alpha(range = c(0.15, 0.65), guide = "none") +  
  ggtitle("Elevation with hillshade") +
  coord_quickmap()

```

</div>


### Credits & Licensing

The contents presented here is drawn from a number of sources including: 

* [Data Carpentary](https://datacarpentry.org/) - Licensed under CC-BY 4.0 2018–2019 by The Carpentries

* [Geocomputation in R](https://geocompr.robinlovelace.net/) - Licences under a Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License.


