---
title: "01 Raster Basics"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE

---
<!--
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document outlines some basic operations for rasters. You can follow along running the code or refer to the html version. 


### Load packages and unzip data files

```{r set-up , echo = TRUE, message= FALSE}

library(raster)
library(usethis)

# set up the data location --------------------------------------

## still need to fix this 
#usethis::use_course("https://github.com/bcgov/bcgov-r-geo-workshop/tree/master/data/20191106_Day_2_PM_Raster/data1.zip") 

#usethis::use_course("https://github.com/bcgov/bcgov-r-geo-workshop/data/20191106_Day_2_PM_Raster/data1.zip")
#usethis::use_zip("https://github.com/bcgov/bcgov-r-geo-workshop/data/20191106_Day_2_PM_Raster/data1.zip")



data.dir <- "C:/Training/R_intermediate/data" #change to common location

# see what goodies are in the folder 
list.files(data.dir)

```
We can see files with a variety of file types (tif, zip etc.)

## Load Rasters 

```{r load Raster}
# read in a single raster 
dem <- raster(file.path(data.dir, "DEM.tif"))

```

## Check Raster 

Unlike vectors, Rasters are defined by the location of grid extents. 
We can look at the key information for rasters; dimension, resolution (cell size), extents, crs, min and max. 


```{r, echo = T, include = T}
dem 

res (dem)

```

We can look at the spread of the data (distribution of values)

```{r, echo = TRUE, include = TRUE, message = FALSE}
summary(dem) # Note this throws an error 
 
summary(dem, maxsamp = ncell(dem)) # forces to check all raster 

cellStats(dem, sd)

```

We can do a quick check of the datasets using standard functions 

```{r, explore the data , echo = T, include = TRUE, results = "hide"}

head(dem) 
values(dem)
unique(values(dem))

```


```{r}
plot(dem)

```

We can format NAs within our data sets 

```{r}

dem <- setMinMax(dem)  
NAvalue(dem) = -9999

```

Again check the minimum numbers 

```{r}
minValue(dem)
maxValue(dem)

```

We can define a range of values which we want to keep 

```{r}
# subset to only include the peaks 
dem.peaks <- clamp(dem, lower = 1000, useValues = TRUE )


#or only view the 
dem.peaks  <- reclassify(dem , cbind(0 , 1000 , NA), right=FALSE)

```


## Visualise Rasters 

We can do quick plots of rasters with both interactive and static maps

```{r}
#quick check of the raster 
plot(dem)

# more fancy exploring 
library(rasterVis)
histogram(dem)
plot3D(dem)

library(mapview)
mapview(dem)

# convert to data frame and plot 
library(ggplot2)
dem_df <- as.data.frame(dem, xy = TRUE)
str(dem_df)

# ggspatial # layer_spatial 

ggplot() +
    geom_raster(data = dem_df , aes(x = x, y = y, fill = DEM)) +
    scale_fill_viridis_c() +
    coord_quickmap()

#...... much more time consuming

```

## Check projections, assign and transform projections

Coordinate Reference System (CRS). We can have CRS in three formats (ESPG, Proj4string, WKT) in proj4string the "+" seperates each element (similar to csv is seperated with a comma)

```{r}
dem
crs(dem)
```


# Questions: what units are the data in ? 
# units = m (meters)

#proj=aea +lat_1=50 +lat_2=58.5 +lat_0=45 +lon_0=-126 + x_0=1000000 + y_0=0 
#ellps=GRS80 
#towgs84=0,0,0,0,0,0,0
#units=m 
#no_defs 

# we can check the information on our ESPG:3005
CRS("+init=epsg:3005")

projection(dem)  # gives us the string version proj4string

# what about projected data ? 
# lets read in another raster 

ndvi <- raster(file.path(data.dir, "NDVI.tif"))
ndvi 

# what is the cell size of the raster? 
# what is the projection? 

projection(ndvi) # what is the crs for this raster - is it the same as the dem layer?

#[1] "+proj=utm +zone=9 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0"

# transform a projection to match our dem 
ndvi.3005 <- projectRaster(ndvi, dem) 

ndvi.3005

summary(ndvi.3005)
#              NDVI
#Min.    -0.2097639
#1st Qu.  0.5884543
#Median   0.6328611
#3rd Qu.  0.6891333
#Max.     0.8716997
#NA's     0.0000000


# What if we want to reclassify our rater and summarise the results 
library(dplyr)

dem.class <- reclassify (dem, c(-Inf, 800, 1,
                                700, 1000, 2, 
                                1000, 1250, 3, 
                                1250, Inf, 4))

plot(dem.class)


dem.class <- as.data.frame(dem.class) %>%
  filter(!is.na(DEM))

dem.class <- dem.class %>%
  group_by(DEM) %>%
  summarise(count = n())


```
