---
title: "01 Raster Basics"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE

---
<!--
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document outlines some basic operations for rasters. You can follow along running the code or refer to the html version. This modual will primarily be usng the `raster` package, however it is worth noting there are a number of other packages currently being developed to handle raster data efficiently. These include;  `stars` and `terra`.  


### 1. Set up 
Firstly we want to set up our work session by reading in required libraries and dataset. We can use `list.files` to take a look at what file types are available.  

```{r set-up , echo = TRUE, message= FALSE, results = "hide"}

library(raster)
library(usethis)

# set up the data location --------------------------------------

## still need to fix this 
#usethis::use_course("https://github.com/bcgov/bcgov-r-geo-workshop/tree/master/data/20191106_Day_2_PM_Raster/data1.zip") 

data.dir <- "C:/Training/R_intermediate/data" #change to common location

# see what goodies are in the folder 
list.files(data.dir)

```

### 2. Loading and checking rasters 

We can read in raster data using the `raster` package. Unlike vectors, Rasters are defined by the location of grid extents, rather than individual vertices. This has some advantages and disadvantages as we will see.  



```{r load Raster, echo = TRUE, message= FALSE, include = T, results = "hide"}
# read in a single raster 

dem <- raster(file.path(data.dir, "DEM.tif"))

```

Once we have read in the raster we can look at the key information; dimension (`dem`), resolution (cell size)(`res`), extents (`extent`), and Co-ordinate Reference System (CRS).

```{r, echo = T, include = T,eval = FALSE}
# dimension 
dim(dem)

#resolution 
res(dem)

# extent
extent(dem)

```

Secondly we can look in detail at the data values. For example we can review the minimum, maximum values and spread of the data (ie. the distribution of values)

```{r, echo = TRUE, include = TRUE, message = FALSE, eval = FALSE}
summary(dem) # Note this throws an error 
 
summary(dem, maxsamp = ncell(dem)) # forces to check all raster 

cellStats(dem, sd)

minValue(dem)

maxValue(dem)

```

As with any `r` object we can do a quick check of the datasets using standard functions (`head`). 

```{r, explore the data , echo = T, include = TRUE, eval = FALSE}

head(dem)           # not very useful

values(dem)         # a little more useful 

unique(values(dem)) # much more useful 

```

We also want to take a look to see if the raster data is what we expect. We can do this with a simple `plot()` function. 

```{r basic plot, results = "hide"}
plot(dem)

```

When reading in rasters it is also important to check how NA's are treated. In this case our `NA`'s are correctly assigned as we see the min and max values do not include NA's. If we read in a raster as see `NA`'s in the min and max values summary we can assign `NA` or reassign to a specific value. For example we can assign all `NA` values a value of 0 or -9999.


```{r, include = TRUE, results = "hide"}
dem <- setMinMax(dem)  

NAvalue(dem) = -9999

```


### 3. Co-ordinate Reference System (CRS)

Coordinate Reference Systems come in three formats (`ESPG`, `proj4string`, `WKT`). Raster data commonly relies on `proj4string` type. In this case the `+` seperates each element (similar to csv is seperated with a comma). Tip: Mapping with leaflet and other web-mapping requires data to be in geographic CRS WGS84 (`EPSG: 4326`).

```{r projections, include = TRUE,  results = "hide"}
dem
crs(dem)

```

The `crs` gives us important information as to what units the data is in, and as a consequence the cell size. For example our `dem` raster is measured in meters (m). This means out resolution is 25m x 25m cell size. 
This is important to understand when looking at area calculations and cell resolution. 

In many cases we use BCAlbers (`EPSG:3005`) as it is an equal area projection with units in meters. 

```{r, include = TRUE, results = "hide"}

# we can check the information on our ESPG:3005
CRS("+init=epsg:3005")

projection(dem)  # gives us the string version proj4string

```


#### Reprojecting Rasters

Reprojecting is different between rasters and vectors. In vectors, each co-ordinate is projected seperately, however as rasters are based on a grid of cells the same size this method is not possible. Raster projections involves creating a new raster of the desired outputs on which the attributes are re-estimated allowing the new pixal to be assigned with the appropriate value. Note the number of columns and rows will be likely to be different from the original raster. 

We can use the `projectRaster()` function. This require a `raster` object and a `crs` argument. This function only accepts `proj4string` definitions of a CRS rather than EPSG codes. 


```{r reproject rasters, include = TRUE, eval = FALSE}
crs(dem)

wgs84 <- "+init=epsg:4326"

dem.wgs84 = projectRaster(dem, crs = wgs84)

plot(dem)
plot(dem.wgs84)

nrow(dem)
nrow(dem.wgs84)

extent(dem)
extent(dem.wgs84)

res(dem)
res(dem.wgs84)

```



### 4. Visualise Rasters and Raster Data

We can do quick visualisations of our rasters using the `plot()`. We can also use other packages to make this view interactive such as `mapview` or `rasterVis`. Another option to take advantage of the `ggplot2` package is to covert our raster into a data.frame and plot the xy variables. This however can be more time consuming as we see from the code below.


```{r raster viz, include = TRUE, eval = FALSE, warning= FALSE,  message = FALSE}
# Option 1: quick check of the raster 
plot(dem)

# Option 2: interactive mapping 
library(mapview)
mapview(dem)

# Option 3: more fancy exploring 
library(rasterVis)
#plot3D(dem)

# Option 4: convert to data frame and plot 
library(ggplot2)
dem_df <- as.data.frame(dem, xy = TRUE)
str(dem_df)

ggplot() +
    geom_raster(data = dem_df , aes(x = x, y = y, fill = DEM)) +
    scale_fill_viridis_c() +
    coord_quickmap()

```

We can also do a plot of the raster values. For example we can create a histogram to look for any anomolies.

```{r histogram, include = TRUE, eval = FALSE}
histogram(dem)

```


### 5. Raster Calculations 

We can do very fast calculation on raster objects as they do not contain explicit spatial projections. For example two rasters with the same extent, resolution and projection we can think of them as two matrices. we can perform calculations based at difference scales; Local (per cell), Focal (neighbourhood), Zonal (neighburhood with iregular size and shape), Global (per raster calcualtions). 


```{r, eval = FALSE}
# Raster Calculations - per cell 
dem + dem 

log(dem)

```

We can reclassify values within the raster. This is useful if we want to bin values into catergories or define a subset of data. Using our `dem` data we can reclassify our values into groups by creating a matrix with defined range and assigned value. We know our minimum value is `r minValue(dem)` and maximum value is `r maxValue(dem)`. 

```{r reclass raster, results = 'hide'}
# reclasss a raster to above or below 1000
rcl <- matrix(c(0, 1000, 1, 1000.0001, 1600, 2), ncol = 3, byrow = TRUE)
rcl 

recl.dem <- reclassify(dem, rcl = rcl)
plot(recl.dem)

# reclassify into 4 groups. 
dem.class <- reclassify (dem, c(-Inf, 800, 1,
                                700, 1000, 2, 
                                1000, 1250, 3, 
                                1250, Inf, 4))

plot(dem.class)

```


Focal operations take into account a cell and its neighbours. Typically these include a 3 x 3 cells. This operation applied an aggregate function to all cells within the neighbourhood with the result being the central cell value. 

We need to define the shape of a moving window with a matrix with corresponding weights. This is commonly summary functions, ie `sum()`, `mean()`, `var()`.

```{r focal , eval = FALSE, include = TRUE}

#Focal operations

r_focal = focal(dem, w = matrix (1,nrow = 3, ncol = 3), fun = min)

plot(dem)
plot(r_focal)

```


Using our dem and functions within the raster package, we can apply focal operations to generate slope and aspect. 


```{r create a hillshade map, eval = T , results="hide"}
# create the terrain layers from a dem based on neighbourhood

slope <- terrain(dem, opt='slope', unit='radians',  neighbors=8)
plot(slope)

aspect <- terrain(dem, opt='aspect', unit='radians',  neighbors=8)
plot(aspect)

```

We can use these outputs to create a hillshade 

```{r hillshade, results = "hide"} 
#Create a hill shade 
hs <- hillShade(slope, aspect, angle=30, direction=270)
plot(hs)

```

Bonus:  We can plot hillshade and dem to create nice feature maps.  

```{rggplot mapping, eval = FALSE}
hs_df <- as.data.frame(hs, xy = TRUE) 
dem_df <- as.data.frame(dem, xy = TRUE)
  
ggplot() +
  geom_raster(data = dem_df , 
              aes(x = x, y = y, 
                  fill = DEM)) + 
  geom_raster(data = hs_df,
              aes(x = x, y = y, alpha = layer)) + 
  scale_fill_gradientn(colours = terrain.colors(10)) +  
  scale_alpha(range = c(0.15, 0.65), guide = "none") +  
  ggtitle("Elevation with hillshade") +
  coord_quickmap()

```


Raster calculator also allows for zonal statistics, for example we can calculating mean, max values per grouping or type. The output is in tabular format. As a demonstration we can use out reclassed raster from above as a quick demonstration. 

```{r zonal stats demo, eval = FALSE}
#zonal calculations 

plot(dem.class) 

dem.zone <- zonal(dem, dem.class, fun= "mean")

dem.zone
```

We will talk more about zonal statistics for remote sensed datasets. 




### Credits & Licensing

The contents presented here is drawn from a number of sources including: 

* [Data Carpentary](https://datacarpentry.org/) - Licensed under CC-BY 4.0 2018–2019 by The Carpentries

* [Geocomputation in R](https://geocompr.robinlovelace.net/) - Licences under a Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License.


