---
title: "02 Raster Manipulation"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Now we have learnt how to read in data and assess the key information and trouble shoot NA and values. 
In normal workflow, howver we will be dealing with multiple rasters which need to be stacked of aligned. 

- What about if we want to stack multiple raster layers and extract values ? 
- How to we match up rasters with different extents and sizes? 


## Aggregate/disaggregate 

```{r}

dem <- raster(file.path(data.dir, "DEM.tif"))

aspect <- raster(file.path(data.dir, "aspect_5m.tif"))


res(aspect) # different extents

# we now know our raster objects are in the same crs
crs(dem)
crs(aspect)

# what about the extent? 
extent(dem)
extent(aspect)

# If we try to stack these they wont stack 
stack(dem,aspect)
```

```{r}
re <- crop(aspect, dem) # 
re <- aggregate(re, fac = 5, fun = mean, expand = TRUE) # Factor 10m * 2.5 25m.
re
stack(dem, re)
res <- resample(re, dem,  fun = mean, expand = TRUE)


# read in slope 
slope <- raster(file.path(data.dir, "slope.tif"))
res(slope)

extent(dem)
extent(slope)

stack(dem, slope)

sl <- crop(slope, dem)
stack(dem, sl, re)



```

# Cropping Raster (by vector)

Lets crop the raster to a specific Bec zone. We can get Bec data from the bcmaps package 

```{r}
library(sf)
library(dplyr)

# first read in our area of interest polygon 

aoi<- st_read(dsn = data.dir,layer = "CanyonCreekBoundary", quiet = TRUE) %>%
   st_transform(3005) 

# read in the bec data 
library(bcmaps)

#available_layers()

bec <- get_layer("bec", class = "sf")

# check CRS

st_crs(bec)
st_crs(aoi)

# Crop to area of interest 
bec_aoi <- st_intersection(bec, aoi)
bec_aoi <- st_cast(bec_aoi, "MULTIPOLYGON")
plot(bec_aoi)


# Extract the SBSmc2 Bec zone and convert to a raster 
bec_aoi <- bec_aoi %>%
  mutate(id = seq(1,length(bec_aoi$MAP_LABEL),1))

bec.r <- fasterize(bec_aoi, dem, field = "id") ###convert to raster

# Extract the SBSmc2 Bec zone and convert to a raster 
sbsmc2 <- bec_aoi %>%
  filter(MAP_LABEL == "SBSmc2") %>%
  mutate(id = 1) 


library(fasterize)

# convert to a raster 

sbsmc2.r <- fasterize(sbsmc2, dem, field = "id") ###convert to raster

plot(dem)
plot(sbsmc2, add = T)

# crop a single raster 
dem.sbsmc2 <- crop(dem, sbsmc2)

# crop a single raster 
dem.sbsmc2 <- crop(dem, sbsmc2.r)



# lets see what is looks like 
plot(dem)
plot(dem.c, add = T, col = "red")


```

# Raster Calculations 

We can do very fast calculation on raster objects as they do not contain explicit spatial projections. For example two rasters with the same extent, resolution and projection we can think of them as two matrices. we can perfomr calculations based at difference scales; Local (per cell), Focal (neighbourhood), Zonal (neighburhood with iregular size and shape), Global (per raster calcualtions). 


```{r}

# Raster Calculations - per cell 

dem + dem 
log(slope)

# for small sized rasters 
rasters.out <- sl - re


# for larger sized rasters (processing intensive)
overlay() # for single layers
calc() # for stacks 

outputRaster <- overlay(dem, sl, fun = mean) 





```



```{r}
minValue(dem)
maxValue(dem)

# 709.151 , 1557.41 
# 709.151 , 1000, 1250, 1557.41 


# reclasss a raster 
rcl <- matrix(c(0, 1000, 1, 1000.0001, 1600, 2), ncol = 3, byrow = TRUE)
recl.dem <- reclassify(dem, rcl = rcl)


```

Focal operations take into account a cell and its neighbours. Typically these include a 3 x 3 cells. This operation applied an aggregate function to all cells within the neighbourhood with the result being the central cell value. 

We need to define the shape of a moving window with a matrix with corresponding weights. This is commonly summary functions, ie sum(), mean(), var()

```{r}

#Focal operations


r_focal = focal(sl, w = matrix (1,nrow = 3, ncol = 3), fun = min)

plot(sl)
plot(r_focal)



# create the terrain layers from a dem based on neighbourhood

slope <- terrain(dem, opt='slope', unit='radians',  neighbors=8)
plot(slope)

aspect <- terrain(dem, opt='aspect', unit='radians',  neighbors=8)
plot(aspect)

# in radians
xr <- terrain(dem, opt=c('slope', 'aspect'), unit='radians',  neighbors=8)
plot(xr)

#Create a hill shade 
hs <- hillShade(slope,aspect, angle=30, direction=270)



```


```{r}
#zonal calculations 

bec.r # bex zones we generated before. 


dem.zone <- zonal(dem, bec.r, fun= "mean")

```



# Mask Value 

For rasters with logical or NA values we can subset using a mask

```{r}
# create raster mask
rmask = dem 
values(rmask) = sample(NA, 36, replace = TRUE)


# spatial subsetting
dem.m <- dem[rmask, drop = FALSE]           # with [ operator
mask(elev, rmask)                   # with mask()
overlay(elev, rmask, fun = "max")   # with overlay

```


## Stack Raster 

So far we have been working with raster layers or single rasters. We can combine these into RasterStacks or RasterBricks. Both of these can hangle multiple layers. 
Raster Bricks are typically correspond to single multispectral satellite image. We will use these later in remote sensing rasters data sets. 

A `RasterStack` is similar in that it contains multiple layers, however all the layers have the same extent and resolution. 

```{r}

# Raster stacks and Bricks 
raster.s <- stack(dem, sl, re)

nlayers(raster.s)
nlayers(dem)

plot(raster.s)

names(raster.s)


cellStats(raster.s, mean)

```




https://geocompr.robinlovelace.net/geometric-operations.html#geo-ras













