---
title: "02 Raster Manipulation"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We have now covered the basics of how to read in raster data, check the basic parameters and do some basic data visualisation. For most analysis however, we typically work with more than one raster and in most cases we are working with raster from different sources and as a consequence are not aligned due to differences in resolution, CRS or extent. We also commonly work with both raster and vector data in an analysis workflow. 

In this section we will work through a typical workflow to preparing multiple rasters, manipulating them to get an common raster alignment, then values from corresponding point dataset.


Within the [raster](https://cran.r-project.org/web/packages/raster/raster.pdf) package there are a number of functions to change the resolution and extent of a raster. 


| Function         |  Details                   | 
|:---------------|:----------------|
| `merge`    |  Combine Raster* objects with different extents (but same origin and resolution)              | |  `mosaic`   |  Combine RasterLayers with different extents and a function for overlap areas         | 
|   `crop`    |  Select a geographic subset of a Raster* object |
|  `extend`  |extend Enlarge a Raster* object
| `trim `   |Trim a Raster* object by removing exterior rows and/or columns that only have NAs |
|  `aggregate`  |  Combine cells of a Raster* object to create larger cells|
|  `disaggregate`  | Subdivide cells |
|  `resample`     |  Warp values to a Raster* object with a different origin or resolution |
| `projectRaster`  |  project values to a raster with a different coordinate reference system |
| `shift`     |   shift Move the location of Raster   |
| `flip`    |  Flip values horizontally or vertically   |
|  `rotate`   |  Rotate values around the date-line (for lon/lat data)  |
|  `t`  |   Transpose a Raster* object   |



## 1. Aggregate & Disaggregate 

Commonly we work with rasters with different resolution. In order to stack rasters on top of each other they need to be aligned with the same resolution and extents. Commonly we will read in rasters with multiple scales, this is important in remote sensing when bands often have a range of scales . For example Sentinel contains 60m and 30m bands. 

```{r read in raster, echo = F, messages = FALSE, results = "hide"}

library(raster)

data.dir <- "C:/Training/R_intermediate/data" #change to common location
data.dir <- "TO BE FILLED IN "

# Get data
unzip("data/20191106_Day_2_PM_Raster/raster_basics.zip", exdir = file.path(data.dir,"raster_basics"))

list.files("data/20191106_Day_2_PM_Raster/")

```

Lets read in a slope layer and build a stack with out `dem` raster. Firstly we need to compare the resolution, extents and crs for each raster. 

```{r  crop example, results = "hide", echo = T}

dem <- raster(file.path(data.dir, "DEM.tif"))

slope <- raster(file.path(data.dir, "slope.tif"))

# check the resolution 
res(slope)
res (dem) 

# check the extents : dont match
extent(dem)
extent(slope)

# try to stack - error 
#stack(dem, slope)

# As extents match we can crop our slope raster to match the Dem 
sl <- crop(slope, dem)

extent(sl)
extent(dem)

```

In this example we used crop to match the extents of the rasters. What about when we have rasters with different resolutions? 

```{r dissagregation part1, include = T, results = "hide"}

aspect <- raster(file.path(data.dir, "aspect.tif"))

res(dem)
res(aspect) # different extents

# we now know our raster objects are in the same crs
crs(dem)
crs(aspect)

# what about the extent? 
extent(dem)
extent(aspect)

# If we try to stack these they wont stack 
#stack(dem, aspect)

```

When we try to stack theses two rasters we get an error message. 
First we need to manipulate the raters to ensure the resolution and extents match. We can use the `aggregate` function to increase the resolution, or `dissagrate` to resample the raster. These functions also allow us to define the method we use to aggregate. This will depend on if our data is catergorical or continuous.

In this case as the data is continuous we can define the function using `mean`, `max`, `min`, etc. 

```{r aggregate to larger resolution , echo = TRUE, results = "hide"}
# We can firstly crop the aspect raster to match the dem
re <- crop(aspect, dem)

# we then aggregate up by a factor of 5. This means search 5 cells 
re <- aggregate(re, fac = 5, fun = mean, expand = TRUE) # Factor 10m * 2.5 25m.

# we can then check the resolution 
re

# now everything looks good we can stack the rasters together
rstack <- stack(dem, re)

rstack 
```

In this example we used `mean` from the surrounding cells to determine the value for our new cell with a larger resolution. The other options include "ngb" nearest neighbour. 


We can also use `resample` to transfer values between non matching raster objects. However it is better practice to explore `aggregate` , `disaggregate`, `crop`, `extend` or `merge` before using `resample`.


```{r}  
# # FIX THIS
# 
# re1 <- crop(aspect, dem)
# res1 <- resample(re1, dem )
# res1
# re
# 
# # note these are difference min values
# 
# res2 <- resample(re1, dem, method = 'bilinear')
# 
# 
# res <- resample(re, dem,  fun = mean, expand = TRUE)
# re
# res1
# res2
# restack <- stack(dem, res)

```

We now have a raster stack which we can work with. 


## 2.  Cropping Raster (by vector)

Another common workflow we use is to crop a raster by a vector layer. This may be a study area or catergorical layers such as a BEC zone. To crop a raster we need a raster and a spatial object to which we want to crop to. 

In this example we can use an area of interest will get the BEC zone data from BC maps and use it as an areas of interest to crop our raster stack. Firstly we will create and aoi using out raster extent then read in and filter out vector layer 

```{r crop by vector, echo = T, results = "hide"}
library(sf)
library(dplyr)
library(bcmaps)

# first read in our area of interest polygon 
aoi <- st_as_sfc(st_bbox(dem)) %>%
  st_transform(3005)

#available_layers()

bec <- get_layer("bec", class = "sf")

# check CRS
st_crs(bec)  
st_crs(aoi)

# Intersect with out area of interest and specify type as "multipolygon"
bec_aoi <- st_intersection(bec, aoi)

bec_aoi <- st_cast(bec_aoi, "MULTIPOLYGON")

plot(bec_aoi)

# Lets saw we are only interested in the SBSmc2 area 
sbsmc2 <- bec_aoi %>%
  filter(MAP_LABEL == "ESSFmcp") %>%
  mutate(id = 1) 

```

We can use our vector to crop the extent of our raster 

```{r}

dem.c <- crop(x = dem, y = as(sbsmc2 , "Spatial"))
plot(dem)
plot(dem.c, col = "red", add = T)


# we can do this on our raster stack 
dem.c <- crop (x = rstack , y = as(sbsmc2, "Spatial"))
plot(dem.c)


```

Another helpful function is to convert vectors into rasters to use as a mask or for analysis which we will show below. 

```{r polygon to raster, echo = T, results = "hide"}

library(fasterize)

# convert to a raster 
sbsmc2.r <- fasterize(sbsmc2, dem, field = "id")   ###convert to 

plot(dem)
plot(sbsmc2, add = T)


```





## Mask Value 

This function created a new raster that has the same values as a defined raster, except for the cells that are NA. These cells will become NA. This works with either another raster or vector data as a sp object. 


```{r mask, echo = T}
# create raster mask
rmask = dem 
#values(rmask) = sample(NA, 36, replace = TRUE)


# spatial subsetting
#dem.m <- dem[rmask, drop = FALSE]           # with [ operator
#mask(elev, rmask)                   # with mask()
#overlay(elev, rmask, fun = "max")   # with overlay

```



https://geocompr.robinlovelace.net/geometric-operations.html#geo-ras




## 1. Raster stacks and Bricks 

So far we have been working with raster layers or single rasters. We can combine these into RasterStacks or RasterBricks. Both of these can hangle multiple layers. 
Raster Bricks are typically correspond to single multispectral satellite image. We will use these later in remote sensing rasters data sets. 

A `RasterStack` is similar in that it contains multiple layers, however all the layers have the same extent and resolution. 


```{r }

# Raster stacks and Bricks 
raster.s <- stack(dem, sl, re)

nlayers(raster.s)
nlayers(dem)

plot(raster.s)

names(raster.s)


cellStats(raster.s, mean)

```


We can define a range of values which we want to keep 

```{r}
# subset to only include the peaks 
dem.peaks <- clamp(dem, lower = 1000, useValues = TRUE )


#or only view the 
dem.peaks  <- reclassify(dem , cbind(0 , 1000 , NA), right=FALSE)


dem.class <- reclassify (dem, c(-Inf, 800, 1,
                                700, 1000, 2, 
                                1000, 1250, 3, 
                                1250, Inf, 4))

plot(dem.class)


```





# extract values from XY from a stack of rasters 




### Credits & Licensing

The contents presented here is drawn from a number of sources including: 

* [Data Carpentary](https://datacarpentry.org/) - Licensed under CC-BY 4.0 2018–2019 by The Carpentries

* [Geocomputation in R](https://geocompr.robinlovelace.net/) - Licences under a Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License.
