---
title: "02 Raster Manipulation"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We have now covered the basics of how to read in raster data, check raster parameters and do some basic data visualisation. For most analysis however, we typically work with more than one raster from different sources. Consequentially rasters are often not aligned due to differences in resolution, CRS or extent. We also commonly work with both raster and vector data in an analysis workflow. 

In this section we will work through a typical workflow to preparing multiple rasters, manipulating them to get an common raster alignment. We will also explore using vector data to extract values for XY point datasets from a stacked raster set. 


Within the [raster](https://cran.r-project.org/web/packages/raster/raster.pdf) package there are a number of functions to manipulate the resolution and extent of a raster. 


| Function         |  Details                   | 
|:---------------|:----------------|
| `merge`    |  Combine Raster* objects with different extents (but same origin and resolution)              | |  `mosaic`   |  Combine RasterLayers with different extents and a function for overlap areas         | 
|   `crop`    |  Select a geographic subset of a Raster* object |
|  `extend`  |extend Enlarge a Raster* object
| `trim `   |Trim a Raster* object by removing exterior rows and/or columns that only have NAs |
|  `aggregate`  |  Combine cells of a Raster* object to create larger cells|
|  `disaggregate`  | Subdivide cells |
|  `resample`     |  Warp values to a Raster* object with a different origin or resolution |
| `projectRaster`  |  project values to a raster with a different coordinate reference system |
| `shift`     |   shift Move the location of Raster   |
| `flip`    |  Flip values horizontally or vertically   |
|  `rotate`   |  Rotate values around the date-line (for lon/lat data)  |
|  `t`  |   Transpose a Raster* object   |



## 1. Raster stacks and Bricks 

An advantage of rasters is that we can stack them together to perform functions on mulitple layers at once. We can use `RasterStacks` or `RasterBricks`. Both of these can hangle multiple layers. It is also important to note other packages are developing methods for storing large raster datasets such as `gdalcubes`. 

Raster Bricks typically correspond to single multispectral satellite image. We will use these later when using remote sensing rasters data sets. 

A `RasterStack` is similar, as it contains multiple layers, however all the layers have the same extent and resolution. 

We can use the previous layers generate in the previous exercise to test these 

```{r set up folders and working dirs, include = F}

library(raster)

# set up the data location --------------------------------------
zip.file <- "../data/20191106_Day_2_PM_Raster/raster_basics.zip"
data.dir <-"../data/20191106_Day_2_PM_Raster/raster_basics"
unzip(zipfile = zip.file, exdir = data.dir)

# see what goodies are in the folder 
file = list.files(data.dir, "DEM.tif", recursive = T, full.names = T)

```

If not saved we can regenerate the layers 

```{r regenerate basic dem layers, include = F}

dem <- raster(file)

slope <- terrain(dem, opt='slope', unit='radians',  neighbors=8)

aspect <- terrain(dem, opt='aspect', unit='radians',  neighbors=8)

```

We can use stack our layers together into a single object and do some basic checks. 

```{r raster stack parameters, include = T, results = 'hide'}
# stack our layers together 
raster.s <- stack(dem, slope, aspect)

# check basic paramters
raster.s

# check the number of layers 
nlayers(raster.s)

# as compared to our dem layer 
nlayers(dem)

# check by plotting 
plot(raster.s)

# check names 
names(raster.s)


cellStats(raster.s, mean)


```



## 2. Aggregate & Disaggregate 

In order to stack rasters on top of each other they need to be aligned with the same resolution and extents. Commonly we will read in rasters with multiple resolutions, this is important in remote sensing when bands often have a range of pixal values. For example Sentinel contains 60m and 30m bands. 

```{r prepare workspace, echo = F, messages = FALSE, results = "hide"}

library(raster)

data.dir <- "C:/Training/R_intermediate/data" #change to common location
#data.dir <- "TO BE FILLED IN "

# Get data
#unzip("data/20191106_Day_2_PM_Raster/raster_basics.zip", exdir = file.path(data.dir,"raster_basics"))

#list.files("data/20191106_Day_2_PM_Raster/")

```

Lets read in a slope layer and build a stack with out `dem` raster. Firstly we need to compare the resolution, extents and crs for each raster. 

```{r  crop example, results = "hide", echo = T}

dem <- raster(file.path(data.dir, "DEM.tif"))

slope <- raster(file.path(data.dir, "slope.tif"))

```

Q. Which parameters should we check to assess if these rasters can align? 

```{r check rasters with different extent, include = FALSE}
# check the crs
crs(slope)
crs(dem)

# check the resolution 
res(slope)
res(dem) 

# check the extents : dont match
extent(slope)
extent(dem)

# try to stack - error 
#stack(dem, slope)

# As extents match we can crop our slope raster to match the Dem 
sl <- crop(slope, dem)

extent(sl)
extent(dem)

stack(dem, sl) # success!

```

In this example we used `crop` to match the extents of the rasters. What about when we have rasters with different resolutions? 

```{r dissagregation part1, include = T, results = "hide"}

aspect <- raster(file.path(data.dir, "aspect.tif"))

res(dem)
res(aspect) # different extents

# we now know our raster objects are in the same crs
crs(dem)
crs(aspect)

# what about the extent? 
extent(dem)
extent(aspect)

# If we try to stack these they wont stack 
#stack(dem, aspect)

```

When we try to stack theses two rasters we get a fairly helpful error message. `Error in compareRaster(x) : different extent`

We need to manipulate these rasters to ensure the resolution and extents match. We can use the `aggregate` function to reduce the resolution (make larger cells), or `dissagrate` to resample the raster (make smaller cells). 
These functions also allow us to define the method we use to aggregate. This will depend primarily on the type of data we have (i.e. catergorical or continuous).

In this case as the data is continuous we can use a number of functions (`mean`, `max`, `min`, etc.) to assign a new value to our pixal. In this example we want the `mean` value from the surrounding cells to determine the value for our new cell with a larger resolution. The other options include "ngb" nearest neighbour. 

```{r aggregate to larger resolution , echo = TRUE, results = "hide"}
 
#?aggregate

# We firstly crop the aspect raster to match the dem
re <- crop(aspect, dem)

# we then aggregate up by a factor of 5. This means our value will be 5 cells in each direction. 
re <- aggregate(re, fact = 5, fun = mean, expand = TRUE) # Factor 5 * 5 =  25m.

# we can then check the resolution 
re

# now everything looks good we can stack the rasters together
rstack <- stack(dem, re)


# we can also add the slope raster we matched to our dem
rstack <- stack(dem, re, sl)

```

Note: We can also use `resample` to transfer values between non matching raster objects. However it is better practice to explore `aggregate` , `disaggregate`, `crop`, `extend` or `merge` before using `resample`.


## 3. Cropping Raster (by vector)

Another common workflow is to crop a raster by a vector layer. This may be a study area (AOI) or catergorical layers such as a BEC zone.

In this example we will use the BEC zone data from `bcmaps` package to define an area of interest. We will then crop our raster to this extent.

Firstly we will create an area of interest using out raster extent. We will use this to filter out vector data. 


```{r crop by vector, echo = T, results = "hide", warning=FALSE}
library(sf)
library(dplyr)
library(bcmaps)

# first read in our area of interest polygon 
aoi <- st_as_sfc(st_bbox(dem)) %>%
  st_transform(3005)

#available_layers()

# lets get the bec layer from the bcmaps package 
bec <- get_layer("bec", class = "sf")

# check CRS
st_crs(bec)  
st_crs(aoi)

# Intersect with out area of interest and specify type as "multipolygon"
bec_aoi <- st_intersection(bec, aoi)

bec_aoi <- st_cast(bec_aoi, "MULTIPOLYGON")

plot(bec_aoi)

# We are only interested in the ESSFmcp type 
sbs <- bec_aoi %>%
  filter(MAP_LABEL == "SBSmc2") %>%
  mutate(id = 1) 

plot(st_geometry(sbs))

```

We now have a vector or area of interest, which we can use to crop our raster or raster stack.  

```{r crop raster by vector, include = TRUE, results = "hide"}

dem.c <- crop(x = dem, y = as(sbs , "Spatial"))

plot(dem)
plot(dem.c, col = "red", add = T)

plot(st_geometry(sbs), add = T)

```

You may notice however in this example we dont exactly have what we want. As raster as based on grids we get the entire grid area, not just that of the vector we wanted. 

Another way to extract the data we want is to `mask` out the areas we dont want. 



## 4. Masking 

This function created a new raster that has the same values as a defined raster, except for the cells that are NA. These cells will become NA. This works with either another raster or vector data as a sp object. 

In this case we can use our bec SBSmc2 vector. We firstly need to convert this to a raster object. We can use the `fasterize` package for quick conversions.  


```{r polygon to raster, echo = T, results = "hide"}

library(fasterize)

# convert to a raster 
sbs.r <- fasterize(sbs, dem, field = "id")   ###convert to 

# plot to see the extents 
plot(dem)
plot(sbs.r, add = T)


# use the SBS raster to mask the values of dem 
dem.m <- mask(dem, sbs.r)

# check the output
plot(dem.m)

# check the spread of the data
summary(dem.m)
summary(dem)

```

We can also perform the `mask` function on raster stacks.  

```{r mask raster stack, include = T, results = "hide"}
rs <- mask(rstack, sbs.r)
plot(rs)

```


## 5. Extract values at points 

Now we have set up our raster layers, we can use our raster stack to extract values at each point. We can use the entire area aoi.  

```{r extract vals, echo = T, include = T, results = "hide"}

rstack <- stack(dem, re, sl)

plot(rstack)

#Lets generate some sample points within our aoi
pts <- st_sample(bec_aoi, size = 25) 


# extract values from single raster 
raster.xy <- extract(dem, st_coordinates(pts))

# extract values from single raster 
raster.xy.s <- extract(rstack, st_coordinates(pts))


```


## 6. Challenge Questions : 
















### Credits & Licensing

The contents presented here is drawn from a number of sources including: 

* [Data Carpentary](https://datacarpentry.org/) - Licensed under CC-BY 4.0 2018–2019 by The Carpentries

* [Geocomputation in R](https://geocompr.robinlovelace.net/) - Licences under a Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License.
