---
title: "02 Raster Manipulation"
author: "G.Perkins (Ministry of Environment and Climate Change Strategy)"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
---

<!--
Copyright 2019 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r generate code only version , eval = FALSE, echo = FALSE}

library(knitr)

purl("20191106_Day_2_PM_Raster/01_Raster_Manipulation.Rmd", 
    output = "20191106_Day_2_PM_Raster/01_Raster_Manipulation_code.R") 

```


We now know how to read in raster data, check parameters and do basic data visualisation and calculations. For most analysis however, we typically work with more than one raster from different sources. Consequentially rasters are often not aligned due to differences in resolution, CRS or extent. We also commonly work with both raster and vector data in an analysis workflow. 

In this section we will work through a typical workflow to preprocess multiple rasters to enable rasters to be stacked together (same extent, resolution, crs). We will also explore using vector data with raster data and extracting values from a stacked raster at XY point locations. 

Within the [raster](https://cran.r-project.org/web/packages/raster/raster.pdf) package there are a number of functions to manipulate the resolution and extent of a raster. 

**Table 1**: Manipulation functions within the `raster` package (replicated from Hijmans, 2011).

| Function         |  Details                   | 
|:---------------|:----------------|
| `merge`    |  Combine Raster* objects with different extents (but same origin and resolution)              | |  `mosaic`   |  Combine RasterLayers with different extents and a function for overlap areas         | 
|   `crop`    |  Select a geographic subset of a Raster* object |
|  `extend`  |extend Enlarge a Raster* object
| `trim `   |Trim a Raster* object by removing exterior rows and/or columns that only have NAs |
|  `aggregate`  |  Combine cells of a Raster* object to create larger cells|
|  `disaggregate`  | Subdivide cells |
|  `resample`     |  Warp values to a Raster* object with a different origin or resolution |
| `projectRaster`  |  project values to a raster with a different coordinate reference system |
| `shift`     |   shift Move the location of Raster   |
| `flip`    |  Flip values horizontally or vertically   |
|  `rotate`   |  Rotate values around the date-line (for lon/lat data)  |
|  `t`  |   Transpose a Raster* object   |


## 1. Raster stacks and Bricks 

An advantage of rasters is that we can stack them together to perform functions on mulitple layers at once. We can use either `RasterStacks` or `RasterBricks`. Note other packages are developing methods for storing large raster datasets such as `gdalcubes`. 

`RasterBricks` typically correspond to single multispectral satellite image. We will use these when dealing with remote sensed datasets. 

A `RasterStack` is similar, as it contains multiple layers, however all the layers have the same extent and resolution. 


```{r set up folders and working dirs, include = F}

library(raster)

data.dir <-"../data/20191106_Day_2_PM_Raster/raster_basics/raster_basics"

file = list.files(data.dir, "DEM.tif", recursive = T, full.names = T)

```


```{r optional local drive,  eval = FALSE, include = FALSE}

# optional: To run locally within Rmarkdown 

#data.dir <-"data/20191106_Day_2_PM_Raster/raster_basics/raster_basics/"

#file = list.files(data.dir, "DEM.tif", recursive = T, full.names = T)

```


We can generate multiple rasters with the same alignment using the terrain functions  (See details in [Bonus Section](https://github.com/bcgov/bcgov-r-geo-workshop/blob/master/20191106_Day_2_PM_Raster/01_Raster_Basics.Rmd)). 
These functions use focal calculations with a DEM `raster` to create slope and aspect rasters. 

```{r regenerate basic dem layers, include = T, results = "hide"}

dem <- raster(file)

rad.slope <- terrain(dem, opt='slope', unit='radians',  neighbors=8)

aspect <- terrain(dem, opt='aspect', unit='radians',  neighbors=8)

```

We can `stack` our layers together into a single object and do some basic checks. 

```{r raster stack parameters, echo= T, results = 'hide'}

# stack our layers together 
raster.s <- stack(dem, rad.slope, aspect)

# check basic paramters
raster.s

# check the number of layers 
nlayers(raster.s)

# as compared to our dem layer 
nlayers(dem)

# check by plotting 
plot(raster.s)

```

<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

#### _Questions:_ 

1) What function would we use to check the names of the rasters within a stack? 

2) Calculate the mean and sd of each raster in the stack?


</div>



## 2. Aggregate & Disaggregate 

In practice, we often obtain raster datasets from multiple sources which are rarely aligned. This means preprocesing is a common requirements in most raster analysis. 

Lets read in another data layer (slope.tiff) and build a stack with our `dem` raster. Firstly we need to compare the resolution, extents and crs for each raster. 

```{r load slope raster, eval = TRUE, echo = T}

file = list.files(data.dir, "slope.tif", recursive = T, full.names = T)

slope <- raster(file)

```

We can check a number of basic paramaters to see if these two rasters can be stacked. 

```{r check rasters with different extent, include = TRUE, results = "hide"}

#  crs
crs(slope)
crs(dem)

# resolution 
res(slope)
res(dem) 

# check the extents 
extent(slope)
extent(dem)

```
We can see the extents dont match for these two rasters. What if we try to stack them anyway?

```{r stack error, eval = FALSE, include = TRUE}

stack(dem, slope)

```

We get a fairly helpful error message! So what can we use to ensure the extents match? 

```{r, eval = TRUE, echo = TRUE, results = "hide"}

# Crop the extents

sl <- crop(slope, dem)

extent(sl)
extent(dem)

stack(dem, sl) # success!

```

In this example we used `crop` to match the extents of the rasters. What about when we have rasters with different resolutions?


<style>
div.blue { background-color:#e6f0ff; border-radius: 5px; padding: 20px;}
</style>
<div class = "blue">

#### _Questions:_ 

1) Read in "aspect.tif" from the data.dir. 

2) Compare the resolution, extent, crs with the dem raster and predict if stack will be successful/ not successful and why? 

* check the Rmd code if you have trouble with this step

</div>

```{r dissagregation part1, eval = TRUE, echo = FALSE, include = FALSE}

file = list.files(data.dir, "aspect.tif", recursive = T, full.names = T)

aspect <- raster(file)

res(dem)
res(aspect) # different extents

# we now know our raster objects are in the same crs
crs(dem)
crs(aspect)

# what about the extent? 
extent(dem)
extent(aspect)

# If we try to stack these they wont stack 
#stack(dem, aspect)

#When we try to stack theses two rasters we get a fairly helpful error message. `Error in compareRaster(x) : different extent`#


```


We need to manipulate these rasters to ensure the resolution and extents match. We can use the `aggregate` function to reduce the resolution (make larger cells), or `dissagrate` to resample the raster (make smaller cells). 
When using the `aggregate` function we can specify how we want our raster values aggregated. This will depend primarily on the type of data we have (i.e. catergorical or continuous).

In this case,  our data is continuous,  we can use a number of functions (`mean`, `max`, `min`, etc.) to assign a new value to our pixal. We will use the `mean` function to calculate the average value for our new cell. Catergorical data uses nearest neighbour ("ngb"). 

```{r aggregate to larger resolution , echo = TRUE, results = "hide"}
# We firstly crop the aspect raster to match the dem

re <- crop(aspect, dem)

# we then aggregate up by a factor of 5. This means our value will be 5 cells in each direction. 
re <- aggregate(re, fact = 5, fun = mean, expand = TRUE) # Factor 5 * 5 =  25m.

# now everything looks good we can stack the rasters together
rstack <- stack(dem, re)

# we can also add the slope raster we matched to our dem
rstack <- stack(dem, re, sl)

```

Note: We can also use `resample` to transfer values between non matching raster objects. However it is better practice to explore `aggregate`, `disaggregate`, `crop`, `extend` or `merge` before using `resample`.


## 3. Cropping Raster (by vector)

Another common workflow is to crop a raster by a vector layer. This may be a study area (AOI) or catergorical layers such as a BEC zone.

In this example we will create an areas of interest from our raster, then extract the BEC zone data from `bcmaps` package and subset to identify the SBSmc2 subzone.

Firstly we will create an area of interest using out raster extent. 

```{r load libraries, include = TRUE, results = "hide", message= FALSE, warning = FALSE}

library(sf)
library(dplyr)
library(bcmaps)

# first read in our area of interest polygon 
aoi <- st_as_sfc(st_bbox(dem)) %>%
  st_transform(3005)

#available_layers()

# lets get the bec layer from the bcmaps package # may take a minute
bec <- get_layer("bec", class = "sf")

# check CRS
st_crs(bec)  
st_crs(aoi)

# Intersect with out area of interest and specify type as "multipolygon"
bec_aoi <- st_intersection(bec, aoi)

bec_aoi <- st_cast(bec_aoi, "MULTIPOLYGON")

plot(bec_aoi)

# We are only interested in the SBSmc2 type : lets assign a number value

sbs <- bec_aoi %>%
  filter(MAP_LABEL == "SBSmc2") %>%
  mutate(id = 1) 

plot(st_geometry(sbs))

```

We now have a vector or area of interest, which we can use to crop our raster or raster stack.  

```{r crop raster by vector, include =TRUE, results = "hide"}

dem.c <- crop(x = dem, y = as(sbs , "Spatial"))

#plot(dem.c)

```

You may notice however in this example we dont exactly have what we want. As raster as based on grids we get the entire grid area, not just that of the vector we wanted. 

```{r plots dem and clip, echo = TRUE, eval = TRUE}
par(mfrow=c(1,2))

plot(dem)

plot(dem.c, col = "red")
plot(st_geometry(sbs), add = T)

dev.off()
```

Another way to extract the data we want is to `mask` out the areas we dont want. 


## 4. Masking Rasters

This function creates a new raster that has the same values as a defined raster, except where the cells are NA. These cells will become `NA`'s. 

In this example we will use our BEC SBSmc2 vector we created above. We firstly convert this to a raster object using the `fasterize` package.   


```{r polygon to raster, echo = T, results = "hide", warning = FALSE, message = FALSE}

library(fasterize)

# convert to a raster 
sbs.r <- fasterize(sbs, dem, field = "id")

# plot to see the extents 
plot(sbs.r)

# use the SBS raster to mask the values of dem 
dem.m <- mask(dem, sbs.r)

# check the output
plot(dem.m)

```

We can also perform the `mask` function on raster stacks.  

```{r mask raster stack, include = T, eval = FALSE}

rs <- mask(rstack, sbs.r)

plot(rs)

```


## 5. Extract values at points 

Now we have set up our raster layers, we can use our raster stack to extract values at each point. We can use the entire area aoi.  

```{r extract vals, echo = T, include = T, results = "hide"}

rstack <- stack(dem, re, sl)

plot(rstack)

#Lets generate some sample points within our aoi
pts <- st_sample(bec_aoi, size = 25) 
pts.xy <- st_coordinates(pts)


plot(dem)
plot(pts, add = TRUE)

# extract values from single raster 
raster.xy <- extract(dem, st_coordinates(pts))

# extract values from single raster 
raster.xy.s <- extract(rstack, st_coordinates(pts))

# add the xy values 
raster.xy.s <- cbind(raster.xy.s, pts.xy)


```

</div>
<style>
div.green{ background-color:#FFFFE0; border-radius: 5px; padding: 20px;}
</style>
<div class = "green">


## _Bonus Section_ 


1) Create a raster stack of the SBSdk BEC zone? 

2) Calculate the mean value for each MAP_LABEL within the dem raster in a table format.



</div>

### Credits & Licensing

The contents presented here is drawn from a number of sources including: 

* [Data Carpentary](https://datacarpentry.org/) - Licensed under CC-BY 4.0 2018–2019 by The Carpentries

* [Geocomputation in R](https://geocompr.robinlovelace.net/) - Licences under a Creative Commons Attribution-NonCommercial-NoDerivatives 4.0 International License.



```{r answers, eval = FALSE, include = FALSE}
# Section 1: 

#1) What function would we use to check the names of the rasters within a stack ? 

names(raster.s)

#2) Calculate the mean and sd of each raster in the stack?  

cellStats(raster.s, mean)
cellStats(raster.s, sd)

# Section 2: 

#### _Questions:_ 

#1) Read in "aspect.tif" from the data.dir ? 

# see inline code above 

#2) Compare the resolution, extent, crs with the dem raster and predict if stack will be successful/ not successful and why? 

# see inline code above 

# Will not align as resolutions and extents are different 

```

